The state of knowledge for marine invertebrate biodiversity in the continental US

### 2017-07-26 manuscript.Rmd
include GP's comments/edits in manuscript

Regional faunistic lists assembled and curated by taxonomic experts provide a comprehensive and detailed account of species that are known to occur in a given geographic area. We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area. We parsed faunistic lists to extract marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1974], leading to `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_species$is_marine, na.rm=TRUE), big.mark=",")` species respectively. Our goals were (1) to compare  the species lists for these geographic areas assembled by experts with the records found in databases; (2) to assess the proportion of species included in the expert lists that have records in public databases.

@@

```{r compare_database_coverage, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right). This plot is generated using the UpSet technique[@Lex2014], which uses a matrix-based layout to represent intersections of sets and their sizes. The rows represent the sets, the columns represent the intersections, and each set that is part of the intersection is marked with a black-filled circle in the cells underneath. Black lines connect the sets that are part of the intersections. (\\#fig:database-coverage)', echo=FALSE}
knitr::include_graphics("figures/database_coverage.pdf")
```

@@

Underscoring the inadequacy of the sampling effort, `r prop_singleton_species`% of the species are known from a single record in iDigBio. Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig. \@ref(fig:idigbio-through-time)). Initial explorations in the late ninetieth century led to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970s when intensive collecting efforts led to an increase in the recording of new taxa (Fig. \@ref(fig:idigbio-through-time)). Since then, the slowing down of collections has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^). The strong correlations between (1) the number of samples collected and the species richness (Fig. \@ref(fig:correlation-sampling)), and (2) the number of specimens collected per year and the number of species recorded for the first time (Fig. \@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and new species. Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions. While less explored than the US, a recent study in Western Australia reported that the collection of 135 locations led to the collection of 805 species of crustaceans and polychaetes, 94.6% and 72% of which were undescribed [@Poore2014], highlighting the potential for discovery in the oceans.

@@

Predicting the geographic range of species, and how they will be impacted by global change has undergone important advances recently [@Guillera-Arroita2017], but remain scarecely used in the marine realm [@Robinson2011]. One limitation is the low number of observations available for each species: it is generally recognized that to obtain reliable results such efforts require to have accurate geographic information for a minimum of 10 specimens but having at least 30 is preferred [@Wisz2008]. From the iDigBio data, we found that `r prop_10_specimens_species`% of species are known from less than 10 records and `r prop_30_specimens_species`% from less than 30 records. This suggests that our understanding of the geographic range of species is limited as most species are known from only a few locations, while efforts to estimate their current and future ranges based on occurrence data are rendered moot by the low number of data points available.

Furthermore, there are important discrepancies in our level of knowledge across phyla. While larger, and accessible groups represent the bulk of the records for marine invertebrates in museum collections, the small organisms, including representatives of the meiofauna are virtually absent. Of the `r n_idigbio_phyla` phyla occurring in iDigBio, `r records_rare_phyla$n_rare_phyla` are represented by less than 500 records, and make up `r records_rare_phyla$prop_rare_phyla`% of the total number of records. Yet, these undersampled taxa are likely to play an important role in the detection of changes in environmental quality due to their size, and physiological vulnerability [@Bik2012]. Additionally, the impact of these changes on higher trophic levels and ecosystems is largely unexplored.

To assess the status of taxonomic expertise in the US, we calculated the proportion of specimens identified at the species level through time. For all taxa, excepts for molluscs, there is a consistent decline, indicating that specimens collected recently remain for the most part unidentified in collections (Fig. \@ref(fig:identification-through-time)). For diverse groups such as annelids or arthropods, the proportion of specimens identified at the species-level is now below 30%. The number of scientists describing new species might be increasing globally [@Costello2013], but taxonomic experts continue to be needed to identify specimens recently collected in the US.

@@

Since its introduction in the early 2000s, DNA barcoding, the sequencing of a common locus across taxa bearing species-specific information, has revitalized taxonomy by facilitating species delineation [@Meyer2005-plos]. More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of species that occur in environmental samples cost effectively [@Leray2015]. This approach opens up new research venues as it becomes possible to estimate the diversity from species-rich samples that were virtually inaccessible before. This is particularly true in marine communities that include multiple phyla and organisms varying in size by orders of magnitude, making identification challenging. However, the utility of the data generated by these projects is enhanced by the possibility of matching DNA sequences obtained to previously sequenced and correctly identified species [@Leray2015]. With decreasing sequencing costs, and the prospect of automating sample collection, these technologies may soon be used routinely to monitor biodiversity continuously [@Bohan2017]. DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions. However, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included in reference databases.

To estimate the proportion of the marine invertebrate diversity for which DNA barcodes are available, we queried BOLD (Barcode of Life Datasystem) for all the species that are listed in the regional lists for the Gulf of Mexico, the Pacific Northwest, as well as all the records for marine invertebrates that occur in the EEZ of the US found in iDigBio. Here, we focused on records based on museum specimens as some of them could potentially be used to complete current DNA barcoding efforts.

The proportion of species that have DNA barcodes available is relatively constant across the better studied phyla at the scale of the EEZ. It varies between 12% for bryozoans to 28% for cnidarians (Fig. @ref(fig:proportion-barcoded)). The proportions for species known to occur in the Gulf of Mexico are usually similar than for the US, while typically a higher fraction of species from the Pacific Northwest are available. In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g. 2.6% for the Antarctic [@Grant2009]). This suggests that even as many additional species have had their DNA barcodes made available since these studies, the US fauna is generally better characterized than other parts of the world.

Using the fraction of known species that have a barcode as a yard stick of the progress made is not sufficient. Many marine invertebrate species are known to be cryptic complexes that are not as widely distributed as currently reported [@Knowlton2000,@Meyer2005-evolution]. Therefore specimens identified under the same species but sampled in distant locations, could represent cryptic species. Geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them). It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced, to capture more accurately their geographic range and the extent of their genetic diversity. This cryptic diversity is also at least partially reflected in the fact that `r stat_bold$n_spp_multi_bin` species out of `r format(stat_bold$n_spp, big.mark=",")` in our dataset include more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]).

```{r proportion-barcoded, fig.cap='Proportion of species known to occur in the US EEZ based on iDigBio records, in the Gulf of Mexico [@Tunnell2009], in the Pacific Northwest[@Kozloff1987], that have publicly available DNA barcodes. Numbers next to the bars indicate the number of species for which DNA barcodes are available. (\\#fig:proportion-barcoded)'}
knitr::include_graphics("figures/proportion_barcoded.pdf")
```

@@

Similarly, a significant fraction of museum records have missing or inaccurate geographic information associated with them. Geographic information might be available in other forms, but needs to be digitized. While tools have been developed to convert names of locales into GPS coordinates on land [@rios2010geolocate], similar approaches are lacking for the sea.

@@

## Material and methods

The list of species for the Gulf of Mexico was obtained by downloading CSV files for each animal phylum from http://e-gulf.org/ [@egulf]. This database includes the species listed in a comprehensive biotic inventory of the Gulf of Mexico [@Felder2009].

The list of species for the Pacific Northwest was downloaded from:  https://www.eopugetsound.org/species/custom-lists/307 (accessed on July 5th, 2017) as a Microsoft Excel spreadsheet. This is a digital version of the species included in [@Kozloff1974].

To search by geographic area, we obtained the boundaries of the Exclusive Economic Zone for the United States from marineregions.org (mrgid:8456) [@mregions]. For the Pacific Northwest, Kozloff did not specify the exact geographic extent covered by his book. We inferred a polygon that encompass part of the Oregon coast to the northern parts of Vancouver island. For the Gulf of Mexico, we used the same geographic area as the one defined in [@Tunnell2009] by modifying manually the eastern boundaries of the shape file for the Gulf of Mexico provided by marineregions.org (mrgid:4288). These geographic areas were split into squares of 0.2°x0.2° which were used to query the web API of iDigBio and OBIS using the R packages ridigbio and robis respectively. The resulting occurrences were then clipped to only include records that fell within the EEZ boundaries.

The species names used in queries or obtained as results were checked against the World Registry of Marine Species (WoRMS). Whenever possible, species names were matched and the current valid names for the species were used. This ensured that the same species listed under different synonyms was only counted once. Species that could not be matched in WoRMS, and species listed as non-marine in WoRMS were not included in subsequent analyses. The R package taxizesoap was used to query the WoRMS database.

To assess the proportion of species for which DNA barcodes are available, we used the current valid names for all records to query the Barcode of Life Datasystem (BOLD) using the R package bold. When no records where found for a species, we looked up for alternative matches using its synonyms as listed in WoRMS.

Figure \@ref(fig:database-coverage) was generated using the R package UpSetR [@Conway2017] based on the UpSet technique [@Lex2014].

>>>>

Regional faunistic lists assembled and curated by taxonomic experts provide a comprehensive and detailed account of species that are known to occur in a given geographic area. We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area. We parsed faunistic lists to extract marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1996], leading to `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_species$is_marine, na.rm=TRUE), big.mark=",")` species respectively. Our goals were (1) to compare  the species lists for these geographic areas assembled by experts with the records found in databases; (2) to assess the proportion of species included in the expert lists that have records in public databases.

@@

```{r compare_database_coverage, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right). This plot is generated using the UpSet technique[@Lex2014;@Conway2017], which uses a matrix-based layout to represent intersections of sets and their sizes. The rows represent the sets, the columns represent the intersections, and each set that is part of the intersection is marked with a black-filled circle in the cells underneath. Black lines connect the sets that are part of the intersections. (\\#fig:database-coverage)', echo=FALSE}
knitr::include_graphics("figures/database_coverage.pdf")
```

@@

Underscoring the inadequacy of the sampling effort, `r prop_singleton_species`% of the species are known from a single record in iDigBio. Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig. \@ref(fig:idigbio-through-time)). Initial explorations in the late ninetieth century led to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970s when intensive collecting efforts led to an increase in the recording of new taxa (Fig. \@ref(fig:idigbio-through-time)). Since then, the slowing down of collections has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^). The strong correlations between (1) the number of samples collected and the species richness (Fig. \@ref(fig:correlation-sampling)), and (2) the number of specimens collected per year and the number of species recorded for the first time (Fig. \@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and new species. <!-- separate issues of sampling is needed, and large-scale sampling is best way of doing it --> Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions. The US marine EEZ is relatively well-known compared to many areas of the world ocean, where studies show that undescribed biodiversity dominates. For example, a recent survey in Western Australia reported that 94.6% of crustaceans and 72% of polychaetes taken at 135 sites were undescribed [@Poore2014].

@@

Environmental conditions are among the primary determinants of species distributions and can thus be used to estimate the response of organisms to climate change. Predictive ecological models forecasting changes in distributions are training on environmental conditions encountered by species in their current distributions. To obtain reliable results, models require at least 10, but ideally 30 or more occurrence records per species [@Wisz2008]. Yet at present, `r prop_10_specimens_species`% of species are represented by $\leq$ 10 records, and `r prop_30_specimens_species`%  by $\leq$ 30 records in iDigBio.

Further biases exist across phyla and body size. Large-bodied and easily acccessible (e.g. dried specimens like shells, corals, echinoderms) are overrepresented in museum collections, compared with small less accessible organisms, such as the meiofauna. Of `r n_idigbio_phyla` phyla represented in iDigBio, `r records_rare_phyla$n_rare_phyla` have <500 records, and make up `r records_rare_phyla$prop_rare_phyla`% of the records. Yet, these undersampled taxa, especially those of small body size, are often the most abundant, and are likely to play an important role in the detection of changes in environmental quality[@Bik2012]. Additionally, the ecological implications of changes in the structure and composition of the communities of these small-bodied organisms, on higher trophic levels and ecosystems are largely unexplored.

The proportion of samples that are identified to species show a marked decline in recent decades in most taxa other than mollusks, demonstrating limiting taxonomic effort and expertise (Fig. \@ref(fig:identification-through-time)). There is a marked break in the central trend arond the 1970's indicating that it now takes decades, on average, to identify samples deposited in collections. For diverse groups such as annelids or arthropods, the proportion of specimens identified to the species-level is now below 30%. The number of scientists describing new species might be increasing globally [@Costello2013], but taxonomic experts continue to be needed to identify specimens recently collected in the US.

@@

Since its introduction in the early 2000s, DNA barcoding, has revitalized taxonomy by facilitating species delineation [@Meyer2005-plos]. More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of specimens in environmental samples (eDNA) cost effectively [@Leray2015], making it possible to estimate diversity and identify taxa from species-rich samples at unprecedented scales. With decreasing sequencing costs, and the prospect of automating sample collection, these technologies may soon be used routinely to monitor biodiversity continuously [@Bohan2017]. However, the utility of the data generated by these projects is limited by the paucity of reliably identified DNA barcode sequences that can serve for matching and identification of eDNA transcripts [@Leray2015]. Thus insights from eDNA technologies depend on the level of completeness and accuracy of comparative, species-level DNA barcode coverage.

<!-- include results earlier? -->

To estimate the proportion of the marine invertebrate diversity for which DNA barcodes are available, we queried BOLD (Barcode of Life Datasystem) for all species listed in the regional lists for the Gulf of Mexico, the Pacific Northwest, as well as all the records for marine invertebrates that occur in the EEZ of the US found in iDigBio.

<!-- add overall mean number before getting into details -->
The proportion of species within the US EEZ that have DNA barcodes varies from 12% for bryozoans to 28% for cnidarians (Fig. \@ref(fig:proportion-barcoded)). The GOM has comparable coverage, while the PNW is better barcoded with >30% species-level coverage across each of the five most diverse phyla. This is driven in part by the active Canadian barcoding effort in British Columbia. In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g. 2.6% for the Antarctic [@Grant2009]). Thus the US fauna is generally better characterized than other parts of the world.

The above estimates are based on species names, but many traditional species include multiple cryptic species [@Knowlton2000;@Meyer2005-evolution]. For example, `r format_output(100*stat_bold$n_spp_multi_bin/stat_bold$n_spp)`% of `r format(stat_bold$n_spp, big.mark=",")` species analyzed here from BOLD include cryptic species, as denoted by assignment to  more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]). Specimens identified under the same species but sampled from distant locations, are especially prone to representing cryptic species. Geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them). It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced, to capture more accurately their geographic range and the extent of their genetic diversity.

```{r proportion-barcoded, fig.cap='Proportion of species known to occur in the US EEZ based on iDigBio records, in the Gulf of Mexico [@Tunnell2009], in the Pacific Northwest[@Kozloff1996], that have publicly available DNA barcodes. Numbers next to the bars indicate the number of species for which DNA barcodes are available. (\\#fig:proportion-barcoded)'}
knitr::include_graphics("figures/proportion_barcoded.pdf")
```

@@

Similarly, a significant fraction of museum records have missing or inaccurate geographic information associated with them. Geographic information might be available in other forms, but needs to be digitized. While tools have been developed to convert names of locales into GPS coordinates on land [@rios2010geolocate], similar approaches are less-developed for the sea.

@@

## Methods

The list of species for the Gulf of Mexico was obtained  from http://e-gulf.org/ [@egulf]. This database includes the species listed in a comprehensive biotic inventory of the Gulf of Mexico [@Felder2009].

The list of species for the Pacific Northwest was downloaded from:  https://www.eopugetsound.org/species/custom-lists/307 (accessed on July 5th, 2017). This is a digital version of the species included in [@Kozloff1996].

For geographic searches, we obtained the boundaries of the Exclusive Economic Zone for the United States from marineregions.org (mrgid:8456) [@mregions]. For the Pacific Northwest, Kozloff did not specify the exact geographic extent covered by his book. We inferred a polygon that encompass part of the Oregon coast to the northern parts of Vancouver island. For the Gulf of Mexico, we used the same geographic area as the one defined in [@Tunnell2009] by modifying manually the eastern boundaries of the shape file for the Gulf of Mexico provided by marineregions.org (mrgid:4288). These geographic areas were split into 0.2°$*$0.2° quadrats that were used to query the web API of iDigBio and OBIS using the R packages ridigbio [@ridigbio-0.3.5] and robis [@robis-0.1.8] respectively. The resulting occurrences were then clipped to only include records that fell within the EEZ boundaries.

All species names obtained from all sources were checked against the World Registry of Marine Species (WoRMS). Species that could not be matched in WoRMS, and species listed as non-marine in WoRMS were not included in subsequent analyses. For all names that were listed in WoRMS, the current accepted name was capture and used in all analyses. This ensured that the same species listed under different synonyms was only counted once. The R package worrms [@worrms-0.1.0] was used to query the WoRMS database.

To assess the proportion of species for which DNA barcodes are available, we queried the Barcode of Life Datasystem (BOLD) using the R package bold [@bold-0.4.0]. We used the accepted names of species in the other lists, as well as synonyms (as listed in WoRMS) when accepted names did not return any records.
